---
title: "index"
author: "Rosealea Bond"
date: "November 30, 2017"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


## `geom_point()`

```{r scatter}

#data(), 

g <- ggplot(data=mpg, aes(x=displ,y=hwy))
g + geom_point()


```

### I need `color`

```{r scatter w/ color}

g + geom_point(aes(color=class))
# or ggplot(data=mpg, aes(x=displ,y=hwy, color=class)) +
#           geom_point() 
```

### Black and white figure with `shape` = `class`

```{r scatter shape}

ggplot(data=mpg, aes(x=displ,y=hwy, shape=class)) +
  geom_point() 
```

### `size`

```{r scatter shape size}

ggplot(data=mpg, aes(x=displ,y=hwy, size=class)) +
  geom_point() 
```

### `size` and `color`

```{r scatter shape color}

ggplot(data=mpg, aes(x=cty,y=hwy, size=class)) +
  geom_point(alpha=0.2, color="blue") 
```

### Chapter 5 example

```{r ch5}
    ggplot(data = mpg)  + 
      geom_point(aes(x = displ, y = hwy), color = "blue")


```

##Geoms

```{r geom_examples}

ggplot(mpg, aes(x=drv,y=hwy))+
  geom_boxplot()

```

###Geoms Violin

```{r geom_violin}

ggplot(mpg, aes(x=drv,y=hwy))+
  geom_violin()

```

### MultipleGeoms 

```{r geom_examples2}

ggplot(mpg) +
  geom_point(aes(x=displ,y=hwy)) +
  geom_smooth(aes(x=displ,y=hwy))

##redundent so you can simplify code

ggplot(mpg, aes(x=displ,y=hwy)) +
  geom_point() +
  geom_smooth() +
labs(title= "rellationship b/w engine size & MPG",
     x= "Highway MPG", 
     y= "Engine Disp.")+
  theme_bw()
  
```


#Data wrangling with `dplyr`

```{r datawrangling}
#reading datafile from github to r
#read_csv is from the readr package not base r
gapminder <- readr::read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv")

gapminder

## Useful functions to get to know your data

head(gapminder)# first six rows
tail(gapminder)# last six rows

head(gapminder,30)# passs as many lines as you want to see

str(gapminder) #check out the structure of your data

names(gapminder) #header

dim(gapminder)# dimensions

ncol(gapminder)
nrow(gapminder)

c(nrow(gapminder),ncol(gapminder)) #combine with c function

summary(gapminder)#summary statistics

#everything above was operating on the whole gapmnder gataset.

head(gapminder$lifeExp)
```
##Now with `dplyr`

```{r filter.example}

filter(gapminder, country=="Mexico")

x <- filter(gapminder,country=="Sweden")
mean(x$lifeExp)
#or
mean(filter(gapminder,country=="Sweden")$lifeExp)

##Pipe operator

gapminder %>% head()

###now use it with the Sweden example

gapminder %>% 
  filter(country=="Sweden") %>% 
  summarise(mean(lifeExp))   #need summarise since it returns a dataframe now to a value. 

##Select by columns

gapminder %>% 
  select(year, lifeExp) %>% 
  tail()

##Compine what we have learned so far

gapminder %>% 
  filter(country=="Cambodia") %>% 
  select(country,year,pop,gdpPercap) #selects 4 of 6

gapminder %>% 
  filter(country == "Cambodia") %>% 
  select(-continent,-lifeExp) #selects all but two mentioned (also 4 of 6)

## a few more things with filter

gapminder %>% 
  filter(country == "Mexico", 
         year == 2002)

##`mutate` to add columns

gapminder %>% 
  mutate(gdp = pop * gdpPercap)

##add index column
gapminder %>% 
  mutate(index = 1:nrow(gapminder))

gapminder %>% 
  filter(country %in% c("Egypt", "Vietnam")) %>% 
  summarise(max(gdpPercap))

## `group_by()`

gapminder %>% 
  group_by(country) %>% 
  summarize(max_gdpPercap = max(gdpPercap))

## recall year with max variable
gapminder %>% 
  group_by(country) %>% 
  mutate(max_gdpPercap = max(gdpPercap)) %>% 
  filter(max_gdpPercap == gdpPercap) %>% 
  arrange(max_gdpPercap) #ascending order
```

#Day 2

##Finishing with `dplyr`

###Joining Datasets

```{r join}
co2 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")

co2 %>% head()
co2 %>% str()

gap_2007<- gapminder %>% 
  filter(year==2007)

##left join gap 2007 to co2

lj<- left_join(gap_2007,co2, by = "country")

##Right join gap 2007 to co2

rj <- right_join(gap_2007,co2)
```

##Starting `tidyr`

###Example Datasets
```{r tidyr datasets}

AirPassengers # wide format
mpg # tidy format (by each observatio)

gap_wide <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')

gapminder

head(gap_wide)
```

##`gather()`
Goal: Turn gap wide into a long dataset

```{r gather}

gap_long <- gap_wide %>% 
  gather(key = obstype_year, 
         value = obs_value,
         dplyr::starts_with ("pop"), #listing which columns to combine
         dplyr::starts_with("lifeExp"),
         dplyr::starts_with("gdpPercap"))
```

### A better way to ignore columns

```{r gather2}

gap_long <- gap_wide %>% 
  gather(key = obstype_year, 
         value = obs_value,
         -continent, -country)

```

### Seperate out the year

```{r seperate}

gap_long <- gap_wide %>% 
  gather(key = obstype_year, 
         value = obs_value,
         -continent, -country) %>% 
  separate(obstype_year, into = c("obs_type", "year"),
           sep = "_", convert = T)

head(gap_long)
```

###Plot Stuff

Long data format
```{r adding plots}

life_df <- gap_long %>% 
  filter(obs_type == "lifeExp",
         continent == "Americas")

head(life_df)

ggplot(data = life_df, aes(x = year, y = obs_value, color = country))+
  geom_line()

```

###Practise 7.7

```{r play}

s <- gap_long %>% 
  filter(obs_type == "lifeExp", 
         year >= 1982 & year <= 2007) %>% 
  group_by(continent, year) %>% 
  summarise(mean_cont = mean(obs_value))

ggplot(s, aes(x = year, y = mean_cont, color = continent)) +
  geom_line()+
  labs(title="Life Expectancy",
       x = "Year",
       y = "Age [y]",
       color = "Continent")+
```

##`spread()`

long to wide (to make it look like the data we started with)

```{r spread}

gap_tolong <- gap_long %>% 
  spread(key=obs_type, 
        value = obs_value)
```

#Programing in R




